---
title: "Feasibilty_Example_wo_account"
author: "Chiara Huwiler"
date: "2023-11-07"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Example for showing feasibility

### Setting up

```{r, eval=FALSE}
# Set up
install.packages("rgbif")
install.packages("dplyr")
install.packages("ggplot2")
install.packages("tidyr")
install.packages("stringr")
install.packages("lubridate")
install.packages("countrycode")
install.packages("rnaturalearth")
install.packages("rnaturalearthdata")

```

```{r, eval=FALSE}
# Load packages
library(rgbif)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(lubridate)
library(countrycode)
library(rnaturalearth)
library(rnaturalearthdata)

# Load Functions
source("./feasibility_example_functions.R")

```

### Data download

The data of tomato with the taxonomy Solanum Lyopersicum L has been
downloaded through the package[@rgbif]. With the code visible in the
following chunk. As the data is already downloaded this chunk doesn't
need to be run and is only for showing how the data has been accessed.

```{r, eval=FALSE}
# Find taxon key
name_backbone("Solanum lycopersicum")

taxonKey <- name_backbone("Solanum lycopersicum")$usageKey
occ_search(taxonKey = taxonKey)

# Downloading data
tomato_data <- occ_download(
  pred("taxonKey", 2930137),
  pred("hasGeospatialIssue", FALSE),
  pred("hasCoordinate", TRUE),
  pred("occurrenceStatus", "PRESENT"),
  pred_not(pred_in("basisOfRecord", c("FOSSIL_SPECIMEN", "LIVING_SPECIMEN"))),
  pred_or(
    pred_lt("coordinateUncertaintyInMeters", 10000),
    pred_isnull("coordinateUncertaintyInMeters")
  ),
  format = "SIMPLE_CSV"
)

# retriving data
 tomato_d <- occ_download_get('0029248-231002084531237') %>%
    occ_download_import()

# saving data
 write.csv(tomato_d, "tomato_d1.csv")
```

### Preparation of the Data

Reducing the dataset to the needed data

```{r}
tomato_d2 <- read.csv("tomato_d1.csv", sep = ",")

# Removing not needed columns
# tomato_d2 <- tomato_d2 %>%
 # select(scientificName, countryCode, year)  

```

Handling missing Data

```{r}
# Deleting all rows with missing scientific name, year or country code
tomato_d2 <- tomato_d2 %>%
  filter(!is.na(scientificName))
tomato_d2 <- tomato_d2 %>%
  filter(!is.na(year))
tomato_d2 <- tomato_d2 %>%
  filter(!is.na(countryCode))


```

### Finding earliest apperance

```{r}
# Finding earliest appearance on a continent for a specific species
source("feasibility_example_functions.R")

# Creating an empty dataframe
result_df <- data.frame(Continent = character(0), Species = character(0), EarliestAppearance = numeric(0))


# Defining continent names and corresponding country codes
continents <- list(
  "Europe" = c("AL", "AD", "AT", "BY", "BE", "BA", "BG", "HR", "CY", "CZ", "DK", "EE", "FI", "FR", "DE", "GR", "HU", "IS", "IE", "IT", "XK", "LV", "LI", "LT", "LU", "MK", "MT", "MD", "MC", "ME", "NL", "NO", "PL", "PT", "RO", "RU", "SM", "RS", "SK", "SI", "ES", "SE", "CH", "TR", "UA", "GB", "VA"),
  "North America" = c("CA", "MX", "US"),
  "South America" = c("AR", "BO", "BR", "CL", "CO", "EC", "GY", "PY", "PE", "SR", "UY", "VE"),
  "Asia" = c("AF", "AM", "AZ", "BH", "BD", "BT", "BN", "KH", "CN", "GE", "IN", "ID", "IR", "IQ", "IL", "JP", "JO", "KZ", "KW", "KG", "LA", "LB", "MY", "MV", "MN", "MM", "NP", "KP", "OM", "PK", "PS", "PH", "QA", "SA", "SG", "KR", "LK", "SY", "TW", "TJ", "TH", "TL", "TM", "AE", "UZ", "VN", "YE"),
  "Africa" = c("DZ", "AO", "BJ", "BW", "BF", "BI", "CV", "CM", "CF", "TD", "KM", "CD", "CG", "CI", "DJ", "EG", "GQ", "ER", "ET", "GA", "GM", "GH", "GN", "GW", "KE", "LS", "LR", "LY", "MG", "MW", "ML", "MR", "MU", "MA", "MZ", "NA", "NE", "NG", "RE", "RW", "ST", "SN", "SC", "SL", "SO", "ZA", "SS", "SD", "SZ", "TZ", "TG", "TN", "UG", "ZM", "ZW"), 
  "Oceania" = c("AU", "NZ", "PG", "FJ", "SB", "VU", "WS", "TO", "KI", "TV", "NR", "PW", "FM", "MH")
)

# Defining the species name
species_name <- "Solanum lycopersicum L."

for (continent in names(continents)) {
  continent_data <- tomato_d2 %>%
    filter(countryCode %in% continents[[continent]], scientificName == species_name)
  
  earliest_appearance <- continent_data %>%
    group_by(scientificName) %>%
    summarize(first_appearance = min(year))
  
  if (nrow(earliest_appearance) > 0) {
    result_df <- rbind(result_df, data.frame(Continent = continent, Species = species_name, EarliestAppearance = earliest_appearance$first_appearance))
  }
}

# Save as a csv
write.csv(result_df, file = "earliest_appearance_continents.csv", row.names = FALSE)

print(result_df)
```

Plotting the result
Ordering the plot form earliest to latest appearance year on a continent.
```{r}

# Ordering the data frame by EarliestAppearance in ascending order
result_df <- result_df %>% arrange(EarliestAppearance)

# Getting the order of continents based on the first appearance year
continent_order <- result_df %>%
  group_by(Continent) %>%
  summarize(first_appearance = min(EarliestAppearance)) %>%
  arrange(first_appearance) %>%
  pull(Continent)

# Defining custom order for y axis
custom_order <- continent_order

# Creating a factor variable for Continent with the custom order
result_df$Continent <- factor(result_df$Continent, levels = custom_order)

# Creating a timeline plot
earliest_appearance_timeline <- ggplot(result_df, aes(x = EarliestAppearance, y = factor(Continent, levels = custom_order))) +
     geom_segment(aes(xend = EarliestAppearance, yend = factor(Continent, levels = custom_order)), color = "skyblue") +
     geom_point(size = 3, color = "darkblue") +
     labs(title = "Timeline of Earliest Appearance of Solanum lycopersicum L.", x = "Year", y = "Continent") +
     theme_minimal()

# Saving the plot
ggsave("earliest_appearance_timeline.png", earliest_appearance_timeline, width = 8, height = 6)

print(earliest_appearance_timeline)

```

### Plotting the Data on a Map

Europe and Asia aren't seperated yet on the plot. For the project this
will be altered.

```{r}
# Getting the continent boundaries
world <- ne_countries(scale = "medium", returnclass = "sf")


# Merging data_EA data frame with continent boundaries
merged_data <- merge(world, result_df, by.x = "continent", by.y = "Continent")


# Creating the map
ggplot(merged_data) +
  geom_sf(aes(fill = EarliestAppearance)) +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(title = "Solanum lycopersicum L. earliest appearance by continent") +
  theme_minimal()

ggsave("earliest_appearance_tomato_plot.png", plot = last_plot(), width = 8, height = 6)

```

### Source

```{r}
# Citation of the downloaded dataset
gbif_citation(x= "0029248-231002084531237")
```