---
title: "Data collection"
author: "Chiara Huwiler"
date: "2023-11-15"
output: html_document
---

```{r, eval=FALSE}
# Set up
install.packages("rgbif")
install.packages("dylyr")
install.packages("httr")
install.packages("curl")
```

```{r, eval=FALSE}
# Load packages
library(rgbif)
library(dplyr)
library(httr)
library(curl)
```

### Data download

Downloading the data of the selected Pizza ingredients through the package[@rgbif]. Attention the download is only possible with a GBIF account. Therefore all the dataset are downloaded and saved in this markdown. Those datasets than can be used during the further process as a csv file.

Selected ingredients

-   Wheat (Triticum)

-   Olive (Olea europaea)

-   Corn (Zea mays)

-   Tomato (Solanum lycopersicum)

-   Onion (Allium cepa)

-   Paprika (Capsicum)

-   Basil (Ocimum basilicum)

While downloading the data already gets filter by the following markers (as recommended by [GBIF](https://docs.ropensci.org/rgbif/articles/getting_occurrence_data.html)):

-   Has Geospatial Issue = FALSE for excluding data with possible bad geographical information

-   Has coordinates = TRUE only getting data with valid geographical coordinates

-   Removing of absent records, fossils and living specimens.

-   Exlusion when the coordinate uncertainity is less than 10'000 m or the uncertainty is not specified.

After the download the data gets simplified by reducing the not needed columns. The now remaining columns scientific name, year and country code get checked for missing value. If there is a missing value, the row will be delete, so that only data remains, that can be used.

#### Wheat

```{r}
# Find taxon key
name_backbone("Triticum")

taxonKey <- name_backbone("Triticum")$usageKey
occ_search(taxonKey = taxonKey)

# Downloading data
wheat_data <- occ_download(
  pred("taxonKey", 2706388),
  pred("hasGeospatialIssue", FALSE),
  pred("hasCoordinate", TRUE),
  pred("occurrenceStatus", "PRESENT"),
  pred_not(pred_in("basisOfRecord", c("FOSSIL_SPECIMEN", "LIVING_SPECIMEN"))),
  pred_or(
    pred_lt("coordinateUncertaintyInMeters", 10000),
    pred_isnull("coordinateUncertaintyInMeters")
  ),
  format = "SIMPLE_CSV"
)

# retriving data
 wheat_d <- occ_download_get('0029248-231002084531237') %>%
    occ_download_import()

# saving data
 write.csv(wheat_d, "data_ingredients/wheat_d1.csv")
```

Reducing the dataset to only the needed data

```{r}
# Removing not needed columns
wheat_d <- wheat_d %>%
  select(scientificName, countryCode, year)  

```

Handling missing Data

```{r}
# Deleting all rows with missing scientific name, year or country code
wheat_d <- wheat_d %>%
  filter(!is.na(scientificName))
wheat_d <- wheat_d %>%
  filter(!is.na(year))
wheat_d <- wheat_d %>%
  filter(!is.na(countryCode))


```

Creating CSV data

```{r}
write.csv(wheat_d, "data_ingredients/cleaned/wheat_d_cleaned.csv")
```

#### Olive

```{r}
# Find taxon key
name_backbone("Olea europaea L.")

taxonKey <- name_backbone("Olea europaea")$usageKey
occ_search(taxonKey = taxonKey)

# Downloading data
olive_data <- occ_download(
  pred("taxonKey", 5415040),
  pred("hasGeospatialIssue", FALSE),
  pred("hasCoordinate", TRUE),
  pred("occurrenceStatus", "PRESENT"),
  pred_not(pred_in("basisOfRecord", c("FOSSIL_SPECIMEN", "LIVING_SPECIMEN"))),
  pred_or(
    pred_lt("coordinateUncertaintyInMeters", 10000),
    pred_isnull("coordinateUncertaintyInMeters")
  ),
  format = "SIMPLE_CSV"
)

# retriving data
 olive_d <- occ_download_get('0029248-231002084531237') %>%
    occ_download_import()

# saving data
 write.csv(olive_d, "data_ingredients/olive_d1.csv")
```

Reducing the dataset to only the needed data

```{r}
# Removing not needed columns
olive_d <- olive_d %>%
  select(scientificName, countryCode, year)  

```

Handling missing Data

```{r}
# Deleting all rows with missing scientific name, year or country code
olive_d <- olive_d %>%
  filter(!is.na(scientificName))
olive_d <- olive_d %>%
  filter(!is.na(year))
olive_d <- olive_d %>%
  filter(!is.na(countryCode))


```

Creating CSV data

```{r}
write.csv(olive_d, "data_ingredients/cleaned/olive_d_cleaned.csv") 
```

#### Corn

```{r}
# Find taxon key
name_backbone("Zea mays")

taxonKey <- name_backbone("Olea europaea")$usageKey
occ_search(taxonKey = taxonKey)

# Downloading data
olive_data <- occ_download(
  pred("taxonKey", 5415040),
  pred("hasGeospatialIssue", FALSE),
  pred("hasCoordinate", TRUE),
  pred("occurrenceStatus", "PRESENT"),
  pred_not(pred_in("basisOfRecord", c("FOSSIL_SPECIMEN", "LIVING_SPECIMEN"))),
  pred_or(
    pred_lt("coordinateUncertaintyInMeters", 10000),
    pred_isnull("coordinateUncertaintyInMeters")
  ),
  format = "SIMPLE_CSV"
)

# retriving data
 corn_d <- occ_download_get('0029248-231002084531237') %>%
    occ_download_import()

# saving data
 write.csv(corn_d, "data_ingredients/corn_d1.csv")
```

Reducing the dataset to only the needed data

```{r}
# Removing not needed columns
corn_d <- corn_d %>%
  select(scientificName, countryCode, year)  

```

Handling missing Data

```{r}
# Deleting all rows with missing scientific name, year or country code
corn_d <- corn_d %>%
  filter(!is.na(scientificName))
corn_d <- corn_d %>%
  filter(!is.na(year))
corn_d <- corn_d %>%
  filter(!is.na(countryCode))


```

Creating CSV data

```{r}
write.csv(corn_d, "data_ingredients/cleaned/corn_d_cleaned.csv") 
```

#### Tomato

```{r}
# Find taxon key
name_backbone("Solanum lycopersicum")

taxonKey <- name_backbone("Solanum lycopersicum")$usageKey
occ_search(taxonKey = taxonKey)

# Downloading data
tomato_data <- occ_download(
  pred("taxonKey", 2930137),
  pred("hasGeospatialIssue", FALSE),
  pred("hasCoordinate", TRUE),
  pred("occurrenceStatus", "PRESENT"),
  pred_not(pred_in("basisOfRecord", c("FOSSIL_SPECIMEN", "LIVING_SPECIMEN"))),
  pred_or(
    pred_lt("coordinateUncertaintyInMeters", 10000),
    pred_isnull("coordinateUncertaintyInMeters")
  ),
  format = "SIMPLE_CSV"
)

# retriving data
 tomato_d <- occ_download_get('0029248-231002084531237') %>%
    occ_download_import()

# saving data
 write.csv(tomato_d, "data_ingredients/tomato_d1.csv")
```

Reducing the dataset to only the needed data

```{r}
# Removing not needed columns
tomato_d <- tomato_d %>%
  select(scientificName, countryCode, year)  

```

Handling missing Data

```{r}
# Deleting all rows with missing scientific name, year or country code
tomato_d <- tomato_d %>%
  filter(!is.na(scientificName))
tomato_d <- tomato_d %>%
  filter(!is.na(year))
tomato_d <- tomato_d %>%
  filter(!is.na(countryCode))


```

Creating CSV data

```{r}
write.csv(tomato_d, "data_ingredients/cleaned/tomato_d_cleaned.csv") # Was "tomato_d1" in the feasibility example
```

### Onion

```{r}
# Find taxon key
name_backbone("Allium cepa")

taxonKey <- name_backbone("Allium cepa")$usageKey
occ_search(taxonKey = taxonKey)

# Downloading data
onion_data <- occ_download(
  pred("taxonKey", 2857697),
  pred("hasGeospatialIssue", FALSE),
  pred("hasCoordinate", TRUE),
  pred("occurrenceStatus", "PRESENT"),
  pred_not(pred_in("basisOfRecord", c("FOSSIL_SPECIMEN", "LIVING_SPECIMEN"))),
  pred_or(
    pred_lt("coordinateUncertaintyInMeters", 10000),
    pred_isnull("coordinateUncertaintyInMeters")
  ),
  format = "SIMPLE_CSV"
)

# retriving data
 onion_d <- occ_download_get('0029248-231002084531237') %>%
    occ_download_import()

# saving data
 write.csv(onion_d, "data_ingredients/onion_d1.csv")
```

Reducing the dataset to only the needed data

```{r}
# Removing not needed columns
onion_d <- onion_d %>%
  select(scientificName, countryCode, year)  

```

Handling missing Data

```{r}
# Deleting all rows with missing scientific name, year or country code
onion_d <- onion_d %>%
  filter(!is.na(scientificName))
onion_d <- onion_d %>%
  filter(!is.na(year))
onion_d <- onion_d %>%
  filter(!is.na(countryCode))

```

Creating CSV data

```{r}
write.csv(onion_d, "data_ingredients/cleaned/onion_d_cleaned.csv")
```

#### Paprika

```{r}
# Find taxon key
name_backbone("Capsicum")

taxonKey <- name_backbone("Capsicum")$usageKey
occ_search(taxonKey = taxonKey)

# Downloading data
paprika_data <- occ_download(
  pred("taxonKey", 2932937),
  pred("hasGeospatialIssue", FALSE),
  pred("hasCoordinate", TRUE),
  pred("occurrenceStatus", "PRESENT"),
  pred_not(pred_in("basisOfRecord", c("FOSSIL_SPECIMEN", "LIVING_SPECIMEN"))),
  pred_or(
    pred_lt("coordinateUncertaintyInMeters", 10000),
    pred_isnull("coordinateUncertaintyInMeters")
  ),
  format = "SIMPLE_CSV"
)

# retriving data
 paprika_d <- occ_download_get('0029248-231002084531237') %>%
    occ_download_import()

# saving data
 write.csv(paprika_d, "data_ingredients/paprika_d1.csv")
```

Reducing the dataset to only the needed data

```{r}
# Removing not needed columns
paprika_d <- paprika_d %>%
  select(scientificName, countryCode, year)  

```

Handling missing Data

```{r}
# Deleting all rows with missing scientific name, year or country code
paprika_d <- paprika_d %>%
  filter(!is.na(scientificName))
paprika_d <- paprika_d %>%
  filter(!is.na(year))
paprika_d <- paprika_d %>%
  filter(!is.na(countryCode))


```

Creating CSV data

```{r}
write.csv(paprika_d, "data_ingredients/cleaned/paprika_d_cleaned.csv")

```

#### Basil ("Ocimum basilicum")

```{r}
# Find taxon key
name_backbone("Ocimum basilicum")

taxonKey <- name_backbone("Ocimum basilicum")$usageKey
occ_search(taxonKey = taxonKey)

# Downloading data
basil_data <- occ_download(
  pred("taxonKey", 2927096),
  pred("hasGeospatialIssue", FALSE),
  pred("hasCoordinate", TRUE),
  pred("occurrenceStatus", "PRESENT"),
  pred_not(pred_in("basisOfRecord", c("FOSSIL_SPECIMEN", "LIVING_SPECIMEN"))),
  pred_or(
    pred_lt("coordinateUncertaintyInMeters", 10000),
    pred_isnull("coordinateUncertaintyInMeters")
  ),
  format = "SIMPLE_CSV"
)

# retriving data
 basil_d <- occ_download_get('0029248-231002084531237') %>%
    occ_download_import()

# saving data
 write.csv(basil_d, "data_ingredients/basil_d1.csv")
```

Reducing the dataset to only the needed data

```{r}
# Removing not needed columns
basil_d <- basil_d %>%
  select(scientificName, countryCode, year)  

```

Handling missing Data

```{r}
# Deleting all rows with missing scientific name, year or country code
basil_d <- basil_d %>%
  filter(!is.na(scientificName))
basil_d <- basil_d %>%
  filter(!is.na(year))
basil_d <- basil_d %>%
  filter(!is.na(countryCode))


```

Creating CSV data

```{r}
write.csv(basil_d, "data_ingredients/cleaned/basil_d_cleaned.csv")

```

# Citation
```{r}
gbif_citation(x= "0029248-231002084531237")
```

