---
title: "Feasibility_Example"
author: "Chiara Huwiler"
date: "2023-11-01"
output: html_document
---

# Example for showing feasibility

### Setting up

```{r}
# Set up
install.packages("rgbif")
install.packages("usethis")
install.packages("dplyr")
install.packages("ggplot2")
install.packages("tidyr")
install.packages("stringr")
install.packages("lubridate")
install.packages("countrycode")
install.packages("rnaturalearth")
install.packages("rnaturalearthdata")

```

```{r}
# Load packages
library(rgbif)
library(usethis)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(lubridate)
library(countrycode)
library(rnaturalearth)
library(rnaturalearthdata)

# gbif downloading is only possible with an account
usethis::edit_r_environ()
```

### Data download

Downloading the data of tomato with the taxonomy Solanum Lyopersicum L. through the package[@rgbif].

```{r}

# Find taxon key
name_backbone("Solanum lycopersicum")

taxonKey <- name_backbone("Solanum lycopersicum")$usageKey
occ_search(taxonKey = taxonKey)

# Downloading data
tomato_data <- occ_download(
  pred("taxonKey", 2930137),
  pred("hasGeospatialIssue", FALSE),
  pred("hasCoordinate", TRUE),
  pred("occurrenceStatus", "PRESENT"),
  pred_not(pred_in("basisOfRecord", c("FOSSIL_SPECIMEN", "LIVING_SPECIMEN"))),
  pred_or(
    pred_lt("coordinateUncertaintyInMeters", 10000),
    pred_isnull("coordinateUncertaintyInMeters")
  ),
  format = "SIMPLE_CSV"
)

# retriving data
 tomato_d <- occ_download_get('0029248-231002084531237') %>%
    occ_download_import()

```

Reducing the dataset to only the needed data

```{r}
# Removing not needed columns
tomato_d <- tomato_d %>%
  select(scientificName, countryCode, year)  

```

Handling missing Data

```{r}
# Deleting all rows with missing scientific name, year or country code
tomato_d <- tomato_d %>%
  filter(!is.na(scientificName))
tomato_d <- tomato_d %>%
  filter(!is.na(year))
tomato_d <- tomato_d %>%
  filter(!is.na(countryCode))


```

Creating CSV data

```{r}
write.csv(tomato_d, "cleaned_data.csv")
```

### Finding earliest apperance

```{r}
# Finding earliest appearance on a continent for a specific species
find_earliest_appearance_species <- function(data, continent_name, country_codes, species_name) {
  continent_data <- data %>%
    filter(countryCode %in% country_codes, scientificName == species_name)
  
  earliest_appearance <- continent_data %>%
    group_by(scientificName) %>%
    summarize(first_appearance = min(year))
  
  if (nrow(earliest_appearance) > 0) {
    cat("Continent:", continent_name, "\n")
    cat("Species:", species_name, "\n")
    cat("Earliest Appearance:", earliest_appearance$first_appearance, "\n\n")
  }
}

# Defining continent names and corresponding country codes
continents <- list(
  "Europe" = c("AL", "AD", "AT", "BY", "BE", "BA", "BG", "HR", "CY", "CZ", "DK", "EE", "FI", "FR", "DE", "GR", "HU", "IS", "IE", "IT", "XK", "LV", "LI", "LT", "LU", "MK", "MT", "MD", "MC", "ME", "NL", "NO", "PL", "PT", "RO", "RU", "SM", "RS", "SK", "SI", "ES", "SE", "CH", "TR", "UA", "GB", "VA"),
  "North America" = c("CA", "MX", "US"),
  "South America" = c("AR", "BO", "BR", "CL", "CO", "EC", "GY", "PY", "PE", "SR", "UY", "VE"),
  "Asia" = c("AF", "AM", "AZ", "BH", "BD", "BT", "BN", "KH", "CN", "GE", "IN", "ID", "IR", "IQ", "IL", "JP", "JO", "KZ", "KW", "KG", "LA", "LB", "MY", "MV", "MN", "MM", "NP", "KP", "OM", "PK", "PS", "PH", "QA", "SA", "SG", "KR", "LK", "SY", "TW", "TJ", "TH", "TL", "TM", "AE", "UZ", "VN", "YE"),
  "Africa" = c("DZ", "AO", "BJ", "BW", "BF", "BI", "CV", "CM", "CF", "TD", "KM", "CD", "CG", "CI", "DJ", "EG", "GQ", "ER", "ET", "GA", "GM", "GH", "GN", "GW", "KE", "LS", "LR", "LY", "MG", "MW", "ML", "MR", "MU", "MA", "MZ", "NA", "NE", "NG", "RE", "RW", "ST", "SN", "SC", "SL", "SO", "ZA", "SS", "SD", "SZ", "TZ", "TG", "TN", "UG", "ZM", "ZW"), 
  "Oceania" = c("AU", "NZ", "PG", "FJ", "SB", "VU", "WS", "TO", "KI", "TV", "NR", "PW", "FM", "MH")
)

# Defining the species name
species_name <- "Solanum lycopersicum L."

# Iterating through continents and finding earliest appearances for the specific species
for (continent in names(continents)) {
  find_earliest_appearance_species(tomato_d, continent, continents[[continent]], species_name)
}
```

```{r}
# Creating a data frame with the continent names, species name, and earliest appearances
data_EA <- data.frame(
  Continent = c("Europe", "North America", "South America", "Asia", "Africa", "Oceania"),
  Species = "Solanum lycopersicum L.",
  EarliestAppearance = c(1737, 1826, 1838, 1834, 1858, 1903)
)

# Reording the levels of the Continent factor variable by EarliestAppearance
data_EA$Continent <- factor(data_EA$Continent, levels = data_EA$Continent[order(data_EA$EarliestAppearance)])

# Creating a timeline plot
ggplot(data_EA, aes(x = as.numeric(EarliestAppearance), y = Continent)) +
  geom_segment(aes(xend = as.numeric(EarliestAppearance), yend = Continent), color = "skyblue") +
  geom_point(size = 3, color = "darkblue") +
  labs(title = "Timeline of Earliest Appearance of Solanum lycopersicum L.", x = "Year", y = "Continent") +
  theme_minimal()

# Saving the plot to a file
ggsave("earliest_appearance_timeline.png", width = 8, height = 6)


```

### Plotting the Data on a Map

Europe and Asia aren't seperated yet on the plot. For the project this will be altered.

```{r}
# Getting the continent boundaries
world <- ne_countries(scale = "medium", returnclass = "sf")


# Merging data_EA data frame with continent boundaries
merged_data <- merge(world, data_EA, by.x = "continent", by.y = "Continent")


# Creating the map
ggplot(merged_data) +
  geom_sf(aes(fill = EarliestAppearance)) +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(title = "Solanum lycopersicum L. earliest appearance by continent") +
  theme_minimal()

ggsave("earliest_appearance_tomato_plot.png", plot = last_plot(), width = 8, height = 6)

```

### Source

```{r}
# Citation of the downloaded dataset
gbif_citation(x= "0029248-231002084531237")
```
